---
title: "Fix Metadata"
output: html_document
date: "2026-04-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Create final WGS and RNA metadata for supp tables and Zenodo

Add in updated C9 and ATXN2 EH calls to WGS metadata
Remove all donor-level info from RNA table
Double check that donor IDs are consistent between WGS and RNA, sex, etc



```{r}
wgs_current <- readxl::read_excel("../../Manuscripts/NYGC_Flagship/Tables/WGS_Metadata_Apr17_2026.xlsx") %>%
  filter(!is.na(`CGND ID`))

rna_current <- readxl::read_excel("../../Manuscripts/NYGC_Flagship/Tables/RNA_Metadata_Apr17_2026.xlsx")

harms_wgs <- readxl::read_excel("Harms_WGS/WGS_Harms_annotation_cleaned.xlsx")

dim(wgs_current)
# 4748 donors
dim(rna_current)
# 2574
```

Fix C9 and ATXN2 calls in WGS

Anna Basile has double checked and these should be the correct ones

```{r}
eh_c9 <- read_tsv("Basile_EH/C9ORF72_summary_table_across_all_projects_GTOnlyForMetaDataTable_nob38.txt") %>% janitor::clean_names()
names(eh_c9)[1] <- "dna_id"
names(eh_c9)[2] <- "c9orf72_genotype"

eh_atxn2 <- read_tsv("Basile_EH/ATXN2_summary_table_across_all_projects_GTOnlyForMetaDataTable_nob38.txt") %>% janitor::clean_names()
names(eh_atxn2)[1] <- "dna_id"
names(eh_atxn2)[2] <- "atxn2_genotype"


get_max <- function(data){
  x <- str_split_fixed(data, "/", 2)
  max <- pmax(as.numeric(x[,1]), as.numeric(x[,2]) )
  return(max)
}

eh_c9$c9orf72_max_call <- get_max(eh_c9$c9orf72_genotype)
eh_atxn2$atxn2_max_call <- get_max(eh_atxn2$atxn2_genotype)
eh_all <- inner_join(eh_c9, eh_atxn2)

```

Harms and Basile match on C9 calls. Good. Added new calls

Check RNA and WGS for concordance

```{r}
rna_current$ExternalSubjectId <- gsub("EC42-C9", "EC42", rna_current$ExternalSubjectId)

rna_donors <- rna_current$ExternalSubjectId
table(rna_donors %in% wgs_current$`Subject ID`)

setdiff(rna_current$ExternalSubjectId, wgs_current$`Subject ID`)

combine <- inner_join(rna_current, wgs_current, by = c("ExternalSubjectId" = "Subject ID") )

overlaps <- intersect(names(rna_current), names(wgs_current))
names(overlaps) <- overlaps

chisq.test(table(combine$`Ethnicity.x`, combine$`Ethnicity.y`))

identical(combine$`Ethnicity.x`, combine$`Ethnicity.y`)

check_overlap <- map_df(overlaps, ~{
  data.frame(
    na_rna = sum(is.na(combine[[paste0(.x, ".x")]]) ),
    na_wgs = sum(is.na(combine[[paste0(.x, ".y")]]) ),
    equal = identical(combine[[paste0(.x, ".x")]], combine[[paste0(.x, ".y")]] ) )
  #broom::tidy(chisq.test(table(combine[[paste0(.x, ".x")]], combine[[paste0(.x, ".y")]])))
}, .id = "variable"
)

check_overlap

```

Seems that RNA metadata is not up to date with WGS metadata on some WGS-derived things, so I should remove those columns from RNA


WGS metadata

add C9 and ATXN2 calls
harmonize names - rna_id, dna_id, donor_id
convert with janitor to all lower case when possible

```{r}
wgs_final <- wgs_current %>%
  janitor::clean_names() %>%
  rename(dna_id = cgnd_id, donor_id = subject_id, dna_platform = platform, dna_library_prep = prep) %>%
  left_join(eh_all, by = "dna_id") %>%
  select(-c9_repeat_size, -expanded_30, -atxn2_repeat_size, -intermediate_30_33) %>%
  select(-data_access_restriction, everything(), data_access_restriction) %>%
  mutate(subject_group_subcategory = ifelse(subject_group_subcategory == "0.0", NA, subject_group_subcategory ))
```


RNA metadata

harmonize names - rna_id, dna_id, donor_id
convert with janitor to all lower case when possible
remove redundant columns in RNA metadata

harmonize tissue names and add tissue codes - maybe


```{r}
tissue_df <- readxl::read_excel("data/QTL/tissue_labels.xlsx")

rna_tissues <- unique(rna_current$`Sample Source`)

rna_current$`Sample Source` %in% tissue_df$region

rna_final <- rna_current %>%
  janitor::clean_names() %>%
  rename(rna_id = internal_id, donor_id = external_subject_id, rna_platform = platform,
         tissue = sample_source, rna_library_prep = prep)

overlap_cols <- intersect(names(wgs_final), names(rna_final))
overlap_cols <- overlap_cols[ overlap_cols != "donor_id"]

# remove additional columns
trim_cols <- c(
  "reported_genomic_mutations_from_sites_not_in_any_way_associated_with_wgs_data_from_nygc",
  "c9_repeat_size",
  "expanded_30",
  "atxn2_repeat_size",
  "intermediate_30_33"
)

rna_final <- select(rna_final, -all_of(overlap_cols), -all_of(trim_cols) )

```


update harms table to match

```{r}
harms_final <- harms_wgs %>%
  janitor::clean_names() %>%
  rename(donor_id = external_subject_id, dna_id = external_sample_id) %>%
  select(-subject_group, -subject_group_subcategory)

harms_final$dna_id2 <- wgs_final$dna_id[match(harms_final$donor_id, wgs_final$donor_id)]

table(is.na(harms_final$dna_id2))
``` 


final checks for interoperability

```{r}
table(rna_final$donor_id %in% wgs_final$donor_id)

table(harms_final$dna_id %in% wgs_final$dna_id)


table(harms_final$donor_id %in% wgs_final$donor_id)
setdiff(harms_final$dna_id, wgs_final$dna_id)


table(harms_final$donor_id %in% rna_final$donor_id)
```



check gene count matrix

```{r}
rna_counts <- read_csv("~/Downloads/count_data_NYGC_ALS_Consortium.csv.gz")

length(unique(names(rna_counts)))
table(names(rna_counts) %in% rna_current$`Internal ID`)

setdiff(names(rna_counts), rna_current$`Internal ID`)
```

Count matrix completely overlaps, hooray.

Write out

```{r}
# weird issue with WGS meta - clinical data has newline characters
wgs_final_fix <- wgs_final %>%
  mutate(across(where(is.character), ~ gsub("[\t\n\r]", " ", .x)))
which(sapply(wgs_final_fix, function(col) any(grepl("\n", col, fixed = TRUE))))

# stupid excel imports C9 and ATXN2 genotypes as dates
wgs_final_fix$c9orf72_genotype <- paste0("EH-",wgs_final_fix$c9orf72_genotype)
wgs_final_fix$atxn2_genotype <- paste0("EH-",wgs_final_fix$atxn2_genotype)


write_tsv(wgs_final_fix, "Metadata/WGS_metadata_apr1726.tsv")
write_tsv(rna_final, "Metadata/RNA_metadata_apr1726.tsv")
write_tsv(harms_final, "Metadata/WGS_annotations_apr176.tsv")

writeLines(names(wgs_final_fix), "Metadata/WGS_columns.txt" )
writeLines(names(rna_final), "Metadata/WGS_columns.txt" )
writeLines(names(harms_final), "Metadata/WGS_columns.txt" )

```

