---
title: "Matt Harms WGS Insights"
author: "Jack Humphrey"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(UpSetR)
```

```{r}
dna_old <- readxl::read_excel("Metadata/ALS Consortium WGS Metadata 022723.xlsx")
rna <- readxl::read_excel("Metadata/ALS Consortium RNAseq Metadata 02212023.xlsx") %>% janitor::clean_names()

dna <- readxl::read_excel("../../Manuscripts/NYGC_Flagship/Tables/ALS_all_supplemental_tables_updated_March_16_2026.xlsx", sheet = 2) %>% janitor::clean_names()

# waiting for Maria to put the CGND DNA IDs back in
#rna <- readxl::read_excel("../../Manuscripts/NYGC_Flagship/Tables/RNA_Supp_Tables/ALS_all_supplemental_tables_updated_March_16_2026.xlsx", sheet = 4) %>% janitor::clean_names()

harms <- readxl::read_excel("Harms_WGS/Harms Path and LP genotypes in dataset 2-22-2026.xlsx") %>% janitor::clean_names()
# fix 5 IDs that got mangled
harms <- harms %>%
  mutate(participant = ifelse( grepl("^HDA", participant), paste0("CGND-HDA-0", gsub("HDA", "", participant) ), participant) )

# add RNA 
harms$external_subject_id <- dna$subject_id[match(harms$participant, dna$cgnd_id)]
harms$rna <- harms$external_subject_id %in% rna$external_subject_id

  
table(harms$participant %in% dna$cgnd_id)
table(harms$participant %in% rna$external_subject_id)
harms$external_subject_id <- dna$subject_id[match(harms$participant, dna$cgnd_id)]


harms_rna <- filter(harms, rna ==TRUE)
#harms_answer <- filter(harms, project == "Answer ALS" )
table(harms_rna$gene1)

harms_rna_c9 <- filter(harms_rna, gene1 == "C9ORF72")
rna$harms_c9 <- rna$external_subject_id %in% harms_rna_c9$external_subject_id

table(rna$sample_source, rna$harms_c9)

```


Fix any bugs or typos

```{r}
#chr1_11022533_G/A - TARDBP mutation but listed as C9
harms$gene2[which(harms$var_id2 == "chr1_11022533_G/A")] <- "TARDBP"

```


ALS-FTD and FTD

Clean up C9 and ATXN2

Using 30 repeats as pathogenic C9 threshold

```{r}
c9_calls <- 
  filter(harms, gene1 == "C9ORF72" | gene2 == "C9ORF72") %>%
  select(participant, protein1, protein2)

c9_calls$c9_1 <- as.numeric(gsub("EH-", "", c9_calls$protein1))
c9_calls$c9_2 <- as.numeric(gsub("EH-", "", c9_calls$protein2))
c9_calls$C9_EH_call <- coalesce(c9_calls$c9_1, c9_calls$c9_2)

harms$C9_EH_call <- c9_calls$C9_EH_call[match(harms$participant, c9_calls$participant)]

#select(harms, -starts_with("x")) %>% View()

harms <- mutate(harms, gene1 = case_when(
  gene1 == "ATXN2" & grepl("Full", comments1) ~ "ATXN2(≥34)",
  gene1 == "ATXN2" & !grepl("Full", comments1) ~ "ATXN2(31-33)",
  gene1 == "C9ORF72" & grepl("Full", comments1) | C9_EH_call >=30 ~ "C9orf72",
  gene1 == "C9ORF72" & !grepl("Full", comments1) ~ "C9orf72(20-29)",
  TRUE ~ gene1
)) %>%
  mutate(gene2 = case_when(
    gene2 == "ATXN2" & grepl("Full", comments2) ~ "ATXN2(≥34)",
    gene2 == "ATXN2" & !grepl("Full", comments2) ~ "ATXN2(31-33)",
    gene2 == "C9ORF72" & (grepl("Full", comments2) | C9_EH_call >=30) ~ "C9orf72",
    gene2 == "C9ORF72" & (!grepl("Full", comments2) | (C9_EH_call >= 20 & C9_EH_call <30) ) ~ "C9orf72(20-29)",
    TRUE ~ gene2
  )) 

harms <- mutate(harms, disease_group = case_when(
  phenotype_details %in% c("Classical/Typical ALS", "Amyotrophic Lateral Sclerosis (ALS)", "ALS Spectrum MND") ~ "ALS",
  phenotype_details %in% c(
    "Frontotemporal Dementia (FTD)", 
    "Behavioral Variant Frontotemporal Dementia (bvFTD)", 
    "Behavioral Variant Frontotemporal Dementia (bvFTD)", 
    "Behavioral Variant Frontotemporal Dementia (bvFTD)",
    "Behavioral Variant Frontotemporal Dementia (bvFTD), Semantic Variant Primary Progressive Aphasia (svPPA)",
    "Frontotemporal Dementia (FTD)",
    "Dementia (NOS)",
    "Non-Fluent/Agrammatic Variant Primary Progressive Aphasia (naPPA)",
    "Dementia (NOS), Progressive Aphasia",
    "Logopenic Variant Primary Progressive Aphasia (lvPPA)",
    "Non-Fluent/Agrammatic Variant Primary Progressive Aphasia (naPPA), Frontotemporal Dementia (FTD)"
  ) ~ "FTD",
  phenotype_details %in% c(
    "Classical/Typical ALS, Frontotemporal Dementia (FTD)",
    "Classical/Typical ALS, Behavioral Variant Frontotemporal Dementia (bvFTD), Progressive Aphasia",
    "Classical/Typical ALS, Alzheimer's Disease (AD), Frontotemporal Dementia (FTD)",
    "Classical/Typical ALS, Frontotemporal Dementia (FTD), Metastatic Carcinoma, Superior Parietal Lobe, Left",
    "Classical/Typical ALS, Frontotemporal Dementia (FTD), Parkinson's Disease (PD)",
    "Classical/Typical ALS, Non-Fluent/Agrammatic Variant Primary Progressive Aphasia (naPPA)",
    "Amyotrophic Lateral Sclerosis (ALS), Dementia (NOS)",
    "Classical/Typical ALS, Alzheimer's Disease (AD)",
    "Classical/Typical ALS, Dementia (NOS)",
    "Classical/Typical ALS, Mild Cognitive Impairment (MCI)",
    "Classical/Typical ALS, Mild Cognitive Impairment (MCI), Myasthenia Gravis",
    "Classical/Typical ALS, Chronic Traumatic Encephalopathy (CTE)",
    "Classical/Typical ALS, Behavioral Variant Frontotemporal Dementia (bvFTD)"
  ) ~ "ALS-FTD",
  grepl("Pre-fALS", phenotype_details) | phenotyp == "Pre-fALS" ~ "Pre-fALS",
  phenotype_details == "Spinal Bulbar Muscular Atrophy (SBMA)" ~ "SBMA",
  phenotype_details %in% c(
    "Primary Lateral Sclerosis (PLS), Frontotemporal Dementia (FTD)",
    "MND (unspecified, non-ALS ALS Spectrum MND), Frontotemporal Dementia (FTD)",
    "Progressive Bulbar Palsy (PBP)",
    "Progressive Bulbar Palsy",
    "Progressive Muscular Atrophy (PMA)",
    "Primary Lateral Sclerosis (PLS)"
  ) ~ "Other MND",
  phenotyp == "Non-Neurological Control" ~ "Control",
  TRUE ~ "Other"
))

## fix oligogenics
harms_oligo <- 
  filter(harms, category_2 %in% c("Oligo", "Driver Oligo", "Bi-allelic", "Homozyogus", "Recessive")
  ) %>%
  split(.$participant) %>%
  map_df( ~{
    g1 <- .x$gene1
    g2 <- .x$gene2
    if(is.na(g2)){ g2 <- g1}
    if(is.na(g1)){g1 <- g2}
    
    if(g1==g2 ){
      .x$gene1 <- paste0(g1, "+", g2)
    }else{
      .x$gene1 <- paste( sort( c(g1, g2) ), collapse = "+" )
    }
    return(.x)
  }
  ) # n = 33
harms$gene <- harms$gene1
harms$gene[ match(harms_oligo$participant, harms$participant) ] <- harms_oligo$gene1

table(harms$disease_group)
table(harms$gene, harms$disease_group)


filter(harms, grepl("C9orf72", gene) ) %>%
  ggplot(aes(x = disease_group, y = C9_EH_call)) + geom_point(aes(colour = gene)) + coord_flip() + scale_y_log10()

filter(harms, grepl("C9orf72", gene), disease_group == "ALS" ) %>%
  mutate(age_at_onset = as.numeric(age_at_onset)) %>%
  mutate(oligo = grepl("\\+", gene)) %>%
  ggplot(aes(x = age_at_onset, y = C9_EH_call)) + geom_point(aes(colour = oligo)) + facet_wrap(~disease_group) +
  facet_wrap(~oligo)


filter(harms, grepl("C9orf72", gene), disease_group == "Other")
```


976 WGS samples annotated by Matt Harms

Breakdown by gene in ALS patients

Multiple categories of genes

1. definite dominant ALS genes (C9, FUS, SOD1, etc) with clnsig1 set as "P" or "P/LP" or "LP"
  Several TARDBP mutations have "Revised to LP" or "Revised to P". One FUS mutation is "Revised from P to VUS" and should be excluded.
 - C9 and ATXN2 should have their repeat thresholds in the plot
 - ATXN2, 45 cases, of which 38 have intermediate expansions (31-33) and 7 have pathogenic expansions (>34)
 
 - some of these are only seen in 1 patient (CHMP2B, MAPT) and are suspect due to links with other diseases

- patient is het for a recessive mutation (SETX, SPG11, SLC52A2, SLC52A3)

Build decision tree
how many ALS patients total?
of which how many had a genetic assessment?

of those, how many had:
- definite dominant variant
- heterozygous for a recessive variant
- likely mimics (HTT, etc)
- uncertain variant

For each type, list the genes, split the C9 and ATXN2 into full and intermediate expansions

```{r}
harms <- mutate(harms, mutation_type = case_when(
  # dominant
  gene1 != "ATXN2(≥34)" &
  category_2 == "Single Variant" &
  !grepl("imic", status1) &
  !grepl("Paget", comments1) &
  !grepl("recessive", comments1) ~ "single dominant variant",
  # recessives
  category_2 == "Single Variant" &
  grepl("recessive", comments1) & 
  !grepl("imic", status1) ~ "recessive variant",
  #mimics
  category_2 == "Single Variant" &
  (grepl("imic", status1) | grepl("Paget", comments1) | gene1 == "ATXN2(≥34)") ~ "potential mimic",
  #oligogenic
  category_2 %in% c("Oligo", "Driver Oligo", "Bi-allelic", "Homozyogus", "Recessive") ~ "oligogenic"
))

```

Check C9 numbers

WGS Metadata is wrong! Known issue, Anna and Maria will fix.

Once I have this I can remake the C9 distribution plot with the new thresholds 20-33 and >34

```{r}
c9 <- filter(harms, grepl("C9orf72", gene) )
# 490 donors have C9 listed by Matt
c9 <- select(c9, -starts_with("x") )
# add prep info
c9$prep <- dna$prep[match(c9$participant, dna$cgnd_id)]
table(c9$prep)

#c9$C9_EH_call <- 
  
# get expansionHunter info
c9_eh_call <- dna$c9_repeat_size
c9_eh_call <- str_split_fixed(c9_eh_call, "/", 2)
dna$c9_eh_call <- pmax( as.numeric(c9_eh_call[,1]), as.numeric(c9_eh_call[,2]))


dna %>%
  ggplot(aes(x = prep, y = c9_eh_call)) + geom_point()

table(dna$c9_eh_call >= 34, dna$prep) # 314
table(dna$c9_eh_call > 20 & dna$c9_eh_call < 34) #13

dna <- mutate(dna, prep_type = case_when(prep %in% c("PCR-Free", "TruSeq PCR-Free") ~ "PCR-free", TRUE ~ "PCR+" ) )
table(dna$prep_type)

table(dna$c9_eh_call == 0)
max(dna$c9_eh_call, na.rm = TRUE)

filter(dna, c9_eh_call > 34) %>% pull(c9_eh_call) %>% median()

# remake C9 distribution plot with new cutoffs
c9_plot <- dna %>%
  filter(!is.na(c9_eh_call) ) %>%
  ggplot(aes(x = 1, y = c9_eh_call )) + 
  geom_violin(trim = TRUE, scale = "count") + 
  geom_jitter(data = filter(dna, c9_eh_call > 20), aes(fill = prep_type), width = 0.2, height = 0, shape = 21) +
  labs(fill = "WGS prep", x = "", y = "Estimated C9orf72 repeat length") +
  scale_y_log10(breaks = c(2, 5, 10, 20, 34, 100, 325, 1198)) +
  geom_hline(yintercept = c(20,34), linetype = 2) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom", 
        legend.direction = "vertical",
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(), 
        axis.text = element_text(colour = "black") ) +
  scale_fill_manual(values = c("darkgray", "red"))
ggsave(plot = c9_plot, filename = "plots/eh_c9_plot_apr26.pdf", width = 2, height = 7 )

c9$c9_eh_call <- dna$c9_eh_call[match(c9$participant, dna$cgnd_id)]

table(c9$c9_eh_call, c9$gene1)

c9$true_eh_call <- as.numeric(gsub("EH-", "", c9$protein1))

c9 %>%
  ggplot(aes(x = gene, y = true_eh_call)) + geom_point()

group_by(c9, gene) %>% slice_min(true_eh_call, n = 1, with_ties = FALSE) %>%
  select(gene, true_eh_call)

```

314 donors found with C9 > 34


```{r}
harms_als <- filter(harms, disease_group == "ALS") # 658
# split ATXN2 and C9orf72 into intermediate and expanded
# combine oligogenic genes

# atxn2 <- filter(harms_als, gene1 == "ATXN2")
# c9 <- filter(harms_als, gene1 == "C9ORF72")
harms_als_def <- filter(harms_als, mutation_type == "single dominant variant")
harms_als_recess <- filter(harms_als, mutation_type ==  "recessive variant")
harms_als_mimic <- filter(harms_als, mutation_type ==  "potential mimic")
harms_als_oligo <- filter(harms_als, mutation_type == "oligogenic")


upset(fromList( list(def = harms_als_def$participant, recess = harms_als_recess$participant, mimic = harms_als_mimic$participant, oligo = harms_als_oligo$participant)) )

```

Plot total gene assignment for WGS and RNA cohort

```{r}
rna_total <- filter(rna, subject_group == "ALS Spectrum MND" & subject_group_subcategory == "Classical/Typical ALS") %>% pull(external_subject_id) %>% unique() %>% length() # 471
rna_gene <- sum(harms_als$rna) # 103

dna_total <- dna %>% filter(subject_group == "ALS Spectrum MND" & subject_group_subcategory == "Classical/Typical ALS") %>% nrow()
dna_gene <- nrow(harms_als)

compare_df <- data.frame(
  type = c("WGS", "RNA-seq"),
  total = c(dna_total, rna_total),
  positive = c(dna_gene, rna_gene)
) 
compare_df
compare_df %>%
  mutate(type_label = paste0(type, "\n(", total, ")")) %>%
  mutate(prop = positive/total) %>%
  mutate(prop = signif(100*prop, 3) ) %>%
  ggplot(aes(x = type_label, y = prop)) + 
  geom_col(aes(fill = type), colour = "black", alpha = 0.5) +
  geom_col(aes(y = prop, fill = type), colour = "black" ) +
  #geom_text(aes(y = total, label = total), colour = "black", nudge_y =5) +
  geom_text(aes(y = prop, label = prop)) + coord_flip() +
  #scale_y_continuous(expand = c(0,0), transform = "log10" ) +
  scale_fill_manual(values = c("red","gray60")) + 
  theme_classic() +
  theme(axis.text.y = element_text( colour = "black"))
  
compare_df

```


Make multiplot for the 4 groups


```{r}
make_gene_plot <- function(data, mytitle = "", log10 = TRUE, clump = 0, guide = TRUE){
  gene_tally <- data %>%
    group_by(gene) %>%
    tally()
  # clump genes with only 1 donor
  if(clump > 0){
  gene_other <- filter(gene_tally, n == clump)
  data$gene <- ifelse(data$gene %in% gene_other$gene, "Other", data$gene)
  }else{
    data$gene <- data$gene
  }
  total_tally <- data %>%
    group_by(gene) %>%
    tally(name = "total") %>%
    arrange(total, desc(gene))
  
  rna_tally <- data %>%
    filter(rna == TRUE) %>%
    group_by(gene) %>%
    tally(name = "rna")
  
  all_tally <- left_join(total_tally, rna_tally, by = "gene") %>%
    #arrange(desc(gene), total ) %>%
    mutate(gene = factor(gene, levels = total_tally$gene)) %>%
    pivot_longer(names_to = "type", values_to = "n", cols = !c(gene))# %>%
    #drop_na()
  
  if(clump >0){
    all_tally <- all_tally %>%
    mutate(gene = forcats::fct_relevel(gene, "Other" ))
  }
  
  p <- all_tally %>%
    ggplot(aes(y = gene, x = log10(n + 1), fill = type )) + 
    geom_col(data = filter(all_tally, type == "total"), colour = "black",width = 0.8 ) +
    #geom_col(aes(x = log10(rna + 1)), fill = "firebrick") +
    geom_col(data = filter(all_tally, type == "rna"), colour = "black", width = 0.8) +
    geom_text(
      data = filter(all_tally, type == "total"),
      aes(label = n, x = log10(n+1), y = gene ), hjust = 0, colour = "black", nudge_x = 0.025) + 
    geom_text(
      data = filter(all_tally, type == "rna"),
      aes(x = log10(n +1), label = n, y = gene), hjust = 1, colour = "white", nudge_x = -0.05) + 
    #scale_x_log10() +
    labs(x = "Donors", y ="", title = mytitle, subtitle = paste0("n = ", nrow(data) )) +
    scale_x_continuous(
      #expand = c(0,0), 
                       expand = expansion(mult = c(0, 0.15)),
                       breaks = log10( c(1, 10, 100, 1000) + 1),
                       labels = c(1, 10, 100, 1000)
    ) +
    theme_classic() +
    theme(axis.text.y = element_text(face = "italic", colour = "black"),
          axis.text.x = element_text(colour = "black"),
          plot.title = element_text(size = 11, colour = "black", margin = margin(0, 0, 2, 15)),
          plot.subtitle = element_text(size = 9, colour = "black", margin = margin(2, 0, 6, 15)),
          plot.title.position = "plot"
          ) +
    scale_fill_manual(values = c("firebrick","dodgerblue4"), labels = c("rna" = "WGS + RNA-seq", "total" = "WGS only") )
  if(guide == FALSE){
    p <- p + guides(fill = "none")
  }
  return(p)
}

```

Make glorious multiplot for main figure

```{r}
multiplot <- (
  ( make_gene_plot(harms_als_def, "Single variant in dominant gene", log10 = TRUE, clump = 1, guide = FALSE) ) / (
  make_gene_plot(harms_als_recess, "Single variant in recessive gene", log10 = TRUE, clump = 1, guide = FALSE ) +
  make_gene_plot(harms_als_mimic, "Potential ALS mimic", log10 = TRUE, clump = 0, guide = FALSE)
  ) + plot_layout(heights = c(14,8) ) |  
    make_gene_plot(harms_als_oligo, "Multiple variants", log10 = TRUE, clump = 0, guide = TRUE)
) +
  plot_layout(widths = c(1,0.2))

ggsave(plot = multiplot, filename = "plots/harms_wgs_annotation_multiplot.pdf", width = 10, height = 8)

multiplot


```

Other diseases for supplement

```{r}
other_plot <- 
make_gene_plot(filter(harms, disease_group == "ALS-FTD"), "ALS-FTD") +

make_gene_plot(filter(harms, disease_group == "FTD"), "FTD") +
make_gene_plot(filter(harms, disease_group == "SBMA"), "SBMA") +
make_gene_plot(filter(harms, disease_group == "Other MND"), "Other MND") +
  plot_layout(heights = c(8,12,2,14), guides = "collect")

other_plot
  
ggsave(plot = other_plot, filename = "plots/harms_wgs_annotation_other_diseases.pdf", width = 6, height = 10)


```

Write out table for supplement


```{r}
t(harms[1,])
harms_supp_df <- select(harms, 
                        ExternalSubjectId = external_subject_id, 
                        ExternalSampleId = participant, 
                        `Disease Group` = disease_group, 
                        `Subject Group` = phenotyp, `Subject Group Subcategory` = phenotype_details, 
                        `RNA-seq` = rna, 
                     `Mutation Group` = mutation_type, `Mutation Type` = category_2, 
                     Mutation = gene, 
                     `Variant ID 1` = var_id1, `Zygosity 1` = zygosity1, `Gene 1` = gene1, `cDNA 1` = c_dna1, `Protein 1` = protein1, `Status 1` = status1, `Clinical significance 1` = clnsig1, `ClinVar allele ID 1` = clin_var_allele_id1, `Comments 1` = comments1,
                     `Variant ID 2` = var_id2, `Zygosity 2` = zygosity2, `Gene 2` = gene2, `cDNA 2` = c_dna2, `Protein 2` = protein2, `Status 2` = status2, `Clinical significance 2` = clnsig2, `ClinVar allele ID 2` = clin_var_allele_id2, `Comments 2` = comments2)

writexl::write_xlsx(harms_supp_df, path = "Harms_WGS/WGS_Harms_annotation_cleaned.xlsx")
```


